### 2022-02-18 tensor_clustering/tensor_clustering_aistats_camera/aistats_camera.tex
aistats

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{caption}
\usepackage[hidelinks]{hyperref}
\usepackage[margin=1in]{geometry}
% \geometry{
%  left=30mm,
%  right=30mm
%  }

\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage[authoryear]{natbib}
\usepackage{url}
\usepackage{enumitem}

@@

\input macros.tex
\newcommand*{\Scale}[2][4]{\scalebox{#1}{$#2$}}%
\setlength\parindent{0pt}
\usepackage[parfill]{parskip}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}

@@

\title{Multiway Spherical Clustering via Degree-Corrected\\ Tensor Block Models\footnote{Part of this paper was presented at 25th International Conference on Artificial Intelligence and Statistics (AISTATS).}}
\date{}
\author{%
%\small
Jiaxin Hu \\
University of Wisconsin -- Madison\\
\texttt{jhu267@wisc.edu} \\
\and
Miaoyan Wang \\
University of Wisconsin -- Madison\\
\texttt{miaoyan.wang@wisc.edu}\\
}

@@

\maketitle
%\doublespacing

\begin{abstract}
We consider the problem of multiway clustering in the presence of unknown degree heterogeneity. Such data problems arise commonly in applications such as recommendation system, neuroimaging, community detection, and hypergraph partitions in social networks. The allowance of degree heterogeneity provides great flexibility in clustering models, but the extra complexity poses significant challenges in both statistics and computation. Here, we develop a degree-corrected tensor block model with estimation accuracy guarantees. We present the phase transition of clustering performance based on the notion of angle separability, and we characterize three signal-to-noise regimes corresponding to different statistical-computational behaviors. In particular, we demonstrate that an intrinsic statistical-to-computational gap emerges only for tensors of order three or greater. Further, we develop an efficient polynomial-time algorithm that provably achieves exact clustering under mild signal conditions. The efficacy of our procedure is demonstrated through two data applications, one on human brain connectome project, and another on Peru Legislation network dataset. 
\end{abstract}

\textbf{Keywords:} tensor clustering, degree correction, statistical-computational efficiency, human brain connectome networks

\section{Introduction}\label{sec:intro}
Multiway arrays have been widely collected in various fields including social networks \citep{anandkumar2014tensor}, neuroscience \citep{wang2017bayesian}, and computer science \citep{koniusz2016sparse}. Tensors effectively represent the multiway data and serve as the foundation in higher-order data analysis. One data example is from multi-tissue multi-individual gene expression study~\citep{wang2019three,hore2016tensor}, where the data tensor consists of expression measurements indexed by (gene, individual, tissue) triplets. Another example is \emph{hypergraph} network~\citep{ghoshdastidar2017uniform,ghoshdastidar2017consistency,ahn2019community,ke2019community} in social science. A $K$-uniform hypergraph can be naturally represented as an order-$K$ tensor, where each entry indicates the presence of $K$-way hyperedge among nodes (a.k.a.\ entities). In both examples, identifying the similarity among tensor entities is important for scientific discovery.

We study the problem of multiway clustering based on a data tensor. The goal of multiway clustering is to identify a checkerboard structure from a noisy data tensor. Figure~\ref{fig:intro} illustrates the noisy tensor and the underlying checkerboard structures discovered by multiway clustering methods. In the hypergraph example, the multiway clustering aims to identify the underlying block partition of nodes based on their higher-order connectivities; therefore, we also refer to the clustering as \emph{higher-order clustering}. The most common model for higher-order clustering is called \emph{tensor block model} (TBM)~\citep{wang2019multiway}, which extends the usual matrix stochastic block model~\citep{abbe2017community} to tensors. The matrix analysis tools, however, are sub-optimal for higher-order clustering. Developing tensor tools for solving block models has received increased interest recently~\citep{ wang2019multiway,chi2020provable,han2020exact}.

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual-specific parameters apart from the community-specific parameters. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. We call the unequal node-specific effects the \emph{degree heterogeneity}. Such degree heterogeneity appears commonly in social networks. Ignoring the degree heterogeneity may seriously mislead the clustering results. For example, regular block model fails to model the member affiliation in the Karate Club network~\citep{bickel2009nonparametric} without addressing degree heterogeneity.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width = .9\textwidth]{intro2_arxiv.pdf}
    \caption{Examples for order-3 tensor block model (TBM) with and without degree correction. Both TBM and dTBM have four communities on each mode, while dTBM allows a richer structure with degree heterogeneity.
    }
    \label{fig:intro}
\end{figure*}

{\bf Our contributions.} The primary goal of this paper is to provide both statistical and computational guarantees for dTBM. Our main contributions are summarized below.
\begin{itemize}[leftmargin=*]

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width = 17cm]{phase.pdf}
    \caption{SNR thresholds for statistical and computational limits in order-$K$ dTBM with dimension $(p,...,p)$ and $K \geq 2$. The SNR gap between statistical possibility and computational efficiency  exists only for tensors with $K \geq 3$. }
    \label{fig:phase_axis}
\end{figure*}

@@

\begin{itemize}[wide]
    \item \textit{Block model for clustering.} Block models such as stochastic block model (SBM) and degree-corrected SBM have been widely used for matrix clustering problems. The theoretical properties and algorithm performance for matrix block models have been well-studied~\citep{gao2018community}; see the review paper~\citep{abbe2017community} and the references therein. However, The tensor counterparts are relatively less understood.

    \item  \textit{Tensor block model.} The (non-degree) tensor block model (TBM) is a higher-order extension of SBM, and its statistical-computational properties are investigated in recent literatures~\citep{wang2019multiway, han2020exact, ghoshdastidar2017consistency}. Extending results from non-degree to degree-corrected model is highly challenging. Our dTBM parameter space is equipped with angle-based similarity and nuisance degree parameters. The extra complexity makes the Cartesian coordinates based analysis~\citep{han2020exact} non-applicable to our setting. Towards this goal, we have developed a new polar coordinates based analysis to control the model complexity. We also develop a new angle-based iteration algorithm to achieve optimal clustering rates \emph{without the need of estimating nuisance degree parameters}.

    \item \textit{Degree-corrected block model.} The hypergraph degree-corrected block model (hDCBM) and its variant have been proposed in the literature~\citep{ke2019community, yuan2018testing}. For this popular model, however, the optimal statistical-computational rates remain an open problem. Our main contribution is to provide a sharp statistical and computational critical phase transition in dTBM literature. In addition, our algorithm results in a faster \emph{exponential} error rate, in contrast to the \emph{polynomial} rate in~\cite{ke2019community}. The original hDCBM~\citep{ke2019community} is designed for binary observations only, and we extend the model to both continuous and binary observations. We believe our results are novel and helpful to the community. See Fig~\ref{fig:phase_axis} for overview of our results.

    \item \textit{Global-to-local algorithm strategy.} Our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{de2000multilinear}, and we develop a weighted higher-order initialization that relaxes the singular-value gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate \emph{provably} achieves the exact clustering within only polynomial-time complexity.

\end{itemize}

\begin{table*}[t]
\resizebox{\textwidth}{!}{%
    \begin{tabular}{c|ccccc}
\hline
    & \cite{gao2018community}&  \cite{han2020exact}&  \cite{ghoshdastidar2017consistency} &\cite{ke2019community} & \textbf{Ours}\\
    \hline
      Allow tensors of arbitrary order & $\times$ & $\surd$ & $\surd$ &$\surd$ & $\surd$  \\
        Allow degree heterogeneity &  $\surd$ & $\times$ & $\surd$ &$\surd$ & $\surd$ \\
        Singular-value gap-free clustering &  $\surd$ & $\surd$ & $\times$ & $\times$ &$\surd$ \\
        Misclustering rate (for order $K^*$)& - & $\exp(-p^{K/2})$ & $p^{-1}$& $p^{-2}$ &  $\exp(-p^{K/2})$\\
        \hline
    \end{tabular}
    }
    \caption{Comparison between previous methods with our method. $^*$We list the result for order-K tensors with $K \geq 3$ and general number of communities $r = \tO(1)$. }\label{tab:comp}
\end{table*}

{\bf Notation.} We use lower-case letters (e.g., $a,b$) for scalars, lower-case boldface letters (e.g., $\ma,\mtheta$) for vectors, upper-case boldface letters (e.g., $\mX,\mY$) for matrices, and calligraphy letters (e.g., $\tX,\tY$) for tensors of order three or greater. We use $\mone_p$ to denote a vector of length $p$ with all entries to be 1. We use $|\cdot|$ for the cardinality of a set and $\ind\{\cdot\}$ for the indicator function. For an integer $p\in\mathbb{N}_{+}$, we use the shorthand $[p]= \offf{1,2,...,p}$. For a length-$p$ vector $\ma$, we use $a(i)\in\mathbb{R}$ to denote the $i$-th entry of $\ma$, and use $\ma_{I}$ to denote the sub-vector by restricting the indices in the set $I\subset [p]$.  We use  $\onorm{\ma}=\sqrt{\sum_{i}a^2(i)}$ to denote the $\ell_2$-norm, $\onorm{\ma}_1=\sum_i |a_i|$ to denote the $\ell_1$ norm of $\ma$. For two vector $\ma, \mb$ of the same dimension, we denote the angle between $\ma, \mb$ by 
\begin{equation}
    \cos \of{\ma, \mb} = \frac{\ang{ \ma, \mb}}{ \onorm{\ma} \onorm{\mb} },
\end{equation}
where $\ang{\ma,\mb}$ is the inner product of two vectors and $\cos \of{\ma, \mb} \in [-1,1]$. We make the convention that $\cos \of{\ma, \mb} = \cos \of{\ma^T, \mb^T}$.

Let $\tY  \in \bbR^{p_1 \times \cdots \times p_K}$ be an order-$K$ $(p_1,...,p_K)$-dimensional tensor. We use $\tY(i_1,\ldots,i_K)$ to denote the $(i_1,\ldots,i_K)$-th entry of $\tY$. The multilinear multiplication of a tensor $\tS\in \bbR^{r_1\times \cdots \times r_K}$ by matrices $\mM_k \in\mathbb{R}^{p_k\times r_k}$ results in an order-$d$ $(p_1,\ldots,p_K)$-dimensional tensor $\tX$, denoted
\[
\tX=\tS \times_1 \mM_1 \times \cdots \times_K \mM_K,
\]
where the entries of $\tX$ are defined by
\begin{align}
    &\tX(i_1,\ldots, i_K)=\sum_{(j_1,\ldots,j_K)}\tS(j_1,\ldots,j_K)\mM_1(i_1,j_1)\cdots \mM_K(i_K,j_K).
\end{align} 
For a matrix $\mY$, we use $\mY_{i:}$ (respectively, $\mY_{:i}$) to denote the $i$-th row (respectively, $i$-th column) of the matrix. Similarly, for an order-3 tensor, we use $\tY_{::i}$ to denote the $i$-th matrix slide of the tensor. We use $\text{Ave}(\cdot)$ to denote the operation of taking averages across elements and $\text{Mat}_k(\cdot)$ to denote the unfolding operation that reshapes the tensor along mode $k$ into a matrix. For a symmetric tensor $\tY\in\mathbb{R}^{p\times \cdots \times p}$, we omit the subscript and use $\text{Mat}(\tY)\in\mathbb{R}^{p\times p^{K-1}}$ to denote the unfolding. For two sequences $\{a_p\}, \{b_p\}$, we denote $a_p\lesssim b_p$ or $a_p=\tO(b_p)$ if $\lim_{p\to\infty}a_p /b_p\leq c$ for some constant $c\geq 0$, $a_p=o(b_p)$ if $\lim_{p\to\infty}a_p/b_p =0$, and $a_p = \Omega(b_p)$ if both $b_p \lesssim a_p$ and $a_p\lesssim b_p$. Throughout the paper, we use the terms ``community'' and ``clusters'' exchangeably.

{\bf Organization.} The rest of this paper is organized as follows. Section~\ref{sec:model} introduces the degree-corrected tensor block model (dTBM) with three motivating examples and presents the identifiability of dTBM under the angle gap condition. We show the phase transition and the existence of statistical-computational gaps for the higher-order dTBM in Section~\ref{sec:limits}. In Section~\ref{sec:alg}, we provide a polynomial-time two-stage algorithm with misclustering rate guarantees. Numerical studies including the simulation, comparison with other methods, and two real dataset analyses are in Sections~\ref{sec:simulation}-\ref{sec:real}. The main technical ideas we develop for addressing main theorems are provided in Section~\ref{sec:mainproof}. Detailed proofs and extra theoretical results are provided in Appendix.

\section{Model formulation and motivations}\label{sec:model}

\subsection{Degree-corrected tensor block model}

Suppose we have an order-$K$ data tensor $\tY \in \bbR^{p \times \cdots \times p}$. For ease of notation, we focus on symmetric tensors in this section; the extension to general asymmetric tensors is provided in Section~\ref{subsec:exten}. Assume there exist $r \geq 2$  disjoint communities among the $p$ nodes. We represent the community assignment by a function $z \colon [p]\mapsto[r]$, where $z(i) = a$ for $i$-th node that belongs to the $a$-th community. Then, $z^{-1}(a)=\{i\in[p]\colon z(i)=a\}$ denotes the set of nodes that belong to the $a$-th community, and $|z^{-1}(a)|$ denotes the number of nodes in the $a$-th community. Let $\mtheta=(\theta(1),\ldots,\theta(p))^T$ denote the degree heterogeneity for $p$ nodes. We consider the order-$K$ dTBM~\citep{ghoshdastidar2017consistency,ke2019community},
\begin{equation}\label{eq:model_margin}
    \tY(i_1,\ldots,i_K) = \tS(z(i_1),\ldots,z(i_K)) \prod_{k = 1}^K\theta_{i_k} + \tE(i_1,\ldots,i_K), 
\end{equation}
\normalsize
where $\tS \in \bbR^{r \times \cdots \times r}$ is an order-$K$ tensor collecting the block means among communities, and $\tE\in\mathbb{R}^{p\times \cdots \times p}$ is a noise tensor consisting of independent zero-mean sub-Gaussian entries with variance bounded by $\sigma^2$. The unknown parameters are $z$, $S$, and $\mtheta$. The dTBM can be equivalently written in a compact form of tensor-matrix product:
\begin{equation}\label{eq:model_tensor}
\bbE\tY = \tS \times_1 \mTheta \mM \times_2 \cdots \times_K  \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\theta(1),...,\theta(p)) \in \bbR^{p \times p}$ is a diagonal matrix, $\mM \in \offf{0,1}^{p \times r}$ is the membership matrix associated with community assignment $z$ such that $\mM(i,j)=\ind\{z(i)=j\}$. By definition, each row of $\mM$ has one copy of 1's and 0's elsewhere. Note that the discrete nature of $\mM$ renders our model~\eqref{eq:model_tensor} more challenging than Tucker decomposition. We call a tensor $\tY$ an $r$-block tensor with degree $\mtheta$ if $\tY$ admits dTBM~\eqref{eq:model_tensor}. The goal of clustering is to estimate $z$ from a single noisy tensor $\tY$. We are particularly interested in the high-dimensional regime where $p$ grows whereas $r=\tO(1)$.

\subsection{Motivating examples} Here, we provide four applications to illustrate the practical necessity of dTBM.

\vspace{0.2cm}
\paragraph{Tensor block model} Consider the model~\eqref{eq:model_tensor}. Let $\theta(i)=1$ for all $ i \in [p]$. The model~\eqref{eq:model_tensor} reduces to the tensor block model, which is widely used in previous clustering algorithms~\citep{wang2019multiway,chi2020provable,han2020exact}. The theoretical results in TBM serve as benchmarks for dTBM.

>>>>

\documentclass[twoside]{article}

%\usepackage{aistats2022}
% If your paper is accepted, change the options for the package
% aistats2022 as follows:
%
\usepackage[accepted]{aistats2022}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

% If you set papersize explicitly, activate the following three lines:
\special{papersize = 8.5in, 11in}
\setlength{\pdfpageheight}{11in}
\setlength{\pdfpagewidth}{8.5in}
% If you use natbib package, activate the following three lines:
\usepackage[round,authoryear]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

% page number margin

%\setlength{\footskip}{10pt}

\usepackage[hidelinks]{hyperref}

\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{url}
\usepackage{enumitem}

@@

\input macros.tex

\newcommand*{\Scale}[2][4]{\scalebox{#1}{$#2$}}%
\setlength\parindent{0pt}
\usepackage[parfill]{parskip}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}

@@

@@

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Multiway Spherical Clustering via Degree-Corrected\\ Tensor Block Models}

\aistatsauthor{ Jiaxin Hu \And Miaoyan Wang }

\aistatsaddress{ University of Wisconsin -- Madison \And  University of Wisconsin -- Madison} ]

\begin{abstract}
We consider the problem of multiway clustering in the presence of unknown degree heterogeneity. Such data problems arise commonly in applications such as recommendation system, neuroimaging, community detection, and hypergraph partitions in social networks. The allowance of degree heterogeneity provides great flexibility in clustering models, but the extra complexity poses significant challenges in both statistics and computation. Here, we develop a degree-corrected tensor block model with estimation accuracy guarantees. We present the phase transition of clustering performance based on the notion of angle separability, and we characterize three signal-to-noise regimes corresponding to different statistical-computational behaviors. In particular, we demonstrate that an intrinsic statistical-to-computational gap emerges only for tensors of order three or greater. Further, we develop an efficient polynomial-time algorithm that provably achieves exact clustering under mild signal conditions. The efficacy of our procedure is demonstrated through both simulations and analyses of Peru Legislation dataset.  
\end{abstract}

\section{INTRODUCTION}\label{sec:intro}
Multiway arrays have been widely collected in various fields including social networks \citep{anandkumar2014tensor}, neuroscience \citep{wang2017bayesian}, and computer science \citep{koniusz2016sparse}. Tensors effectively represent the multiway data and serve as the foundation in higher-order data analysis. One data example is from multi-tissue multi-individual gene expression study~\citep{wang2019three,hore2016tensor}, where the data tensor consists of expression measurements indexed by (gene, individual, tissue) triplets. Another example is \emph{hypergraph} network~\citep{ghoshdastidar2017uniform,ghoshdastidar2017consistency,ahn2019community,ke2019community} in social science. A $K$-uniform hypergraph can be naturally represented as an order-$K$ tensor, where each entry indicates the presence of $K$-way hyperedge among nodes (a.k.a.\ entities). In both examples, identifying the similarity among tensor entities is important for scientific discovery.

We study the problem of multiway clustering based on a data tensor. The goal of multiway clustering is to identify a checkerboard structure from a noisy data tensor. Figure~\ref{fig:intro} illustrates the noisy tensor and the underlying checkerboard structures discovered by multiway clustering methods. In the hypergraph example, the multiway clustering aims to identify the underlying block partition of nodes based on their higher-order connectivities; therefore, we also refer to the clustering as \emph{higher-order clustering}. The most common model for higher-order clustering is called \emph{tensor block model} (TBM)~\citep{wang2019multiway}, which extends the usual matrix stochastic block model~\citep{abbe2017community} to tensors. Matrix analysis tools, however, are sub-optimal for higher-order clustering. Developing tensor tools for solving block models has received increased interest recently~\citep{ wang2019multiway,chi2020provable,han2020exact}.

Classical tensor block model suffers from drawbacks to model real world data in spite of the popularity. The key underlying assumption of block model is that all nodes in the same community are exchangeable; i.e., the nodes have no individual-specific parameters apart from the community-specific parameters. However, the exchangeability assumption is often non-realistic. Each node may contribute to the data variation by its own multiplicative effect. We call the unequal node-specific effects the \emph{degree heterogeneity}. Such degree heterogeneity appears commonly in social networks. Ignoring the degree heterogeneity may seriously mislead the clustering results. For example, regular block model fails to model the member affiliation in Karate Club network~\citep{bickel2009nonparametric} without addressing degree heterogeneity.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width = .9\columnwidth]{intro2_2.pdf}
    \caption{ \small Examples for order-3 TBM with and without degree correction. Both TBM and dTBM have four communities on each mode, while dTBM allows a richer structure with degree heterogeneity.
    }
    \label{fig:intro}
\end{figure}

{\bf Our contributions.} The primary goal of this paper is to provide both statistical and computational guarantees for dTBM. Our main contributions are summarized below.
\begin{itemize}[wide,topsep=-2pt,itemsep=2pt,parsep=4pt]

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width = .9\textwidth]{phase.pdf}
    \caption{ \small SNR thresholds for statistical and computational limits in order-$K$ dTBM with dimension $(p,...,p)$ and $K \geq 2$. The SNR gap between statistical possibility and computational efficiency  exists only for tensors with $K \geq 3$. }
    \label{fig:phase_axis}
\end{figure*}

\thispagestyle{empty}

@@

    \textit{Block model.} Block models such as stochastic block model (SBM) and degree-corrected SBM have been widely used for matrix clustering problems. See the review paper~\citep{abbe2017community} and the references therein.
    %However, The tensor counterparts are relatively less understood.  
    The (non-degree) tensor block model (TBM) is a higher-order extension of SBM, and its statistical-computational properties are investigated in recent literatures~\citep{wang2019multiway, han2020exact, ghoshdastidar2017consistency}. Extending results from non-degree to degree-corrected model is highly challenging. Our dTBM parameter space is equipped with angle-based similarity and nuisance degree parameters. The extra complexity makes the Cartesian coordinates based analysis~\citep{han2020exact} non-applicable to our setting. Towards this goal, we have developed a new polar coordinates based analysis to control the model complexity. We also develop a new angle-based iteration algorithm to achieve optimal clustering rates \emph{without the need of estimating nuisance degree parameters}.

     \textit{Degree-corrected block model.} The hypergraph degree-corrected block model (hDCBM) and its variant have been proposed in the literature~\citep{ke2019community, yuan2018testing}. For this popular model, however, the optimal statistical-computational rates remain an open problem. Our main contribution is to provide a sharp statistical and computational critical phase transition in dTBM literature. In addition, our algorithm results in a faster \emph{exponential} error rate, in contrast to the \emph{polynomial} rate in~\cite{ke2019community}. The original hDCBM~\citep{ke2019community} is designed for binary observations only, and we extend the model to both continuous and binary observations. We believe our results are novel and helpful to the community. See Figure~\ref{fig:phase_axis} for overview of our results.

    \textit{Global-to-local algorithm strategy.} Our methods generalize the recent global-to-local strategy for matrix learning~\citep{gao2018community,chi2019nonconvex,yun2016optimal} to tensors~\citep{han2020exact,ahn2018hypergraph,kim2018stochastic}. Despite the conceptual similarity, we address several fundamental challenges associated with this non-convex, non-continuous problem. We show the insufficiency of the conventional tensor HOSVD~\citep{de2000multilinear}, and we develop a weighted higher-order initialization that relaxes the singular-value gap separation condition. Furthermore, our local iteration leverages the angle-based clustering in order to avoid explicit estimation of degree heterogeneity. Our bounds reveal the interesting interplay between the computational and statistical errors. We show that our final estimate \emph{provably} achieves the exact clustering within only polynomial-time complexity.

\begin{table*}[t]
\resizebox{\textwidth}{!}{%
\centering
    \begin{tabular}{c|ccccc}
\hline
    & \cite{gao2018community}&  \cite{han2020exact}&  \cite{ghoshdastidar2017consistency} &\cite{ke2019community} & \textbf{Ours}\\
    \hline
      Allow tensors of arbitrary order & $\times$ & $\surd$ & $\surd$ &$\surd$ & $\surd$  \\
        Allow degree heterogeneity &  $\surd$ & $\times$ & $\surd$ &$\surd$ & $\surd$ \\
        Singular-value gap-free clustering &  $\surd$ & $\surd$ & $\times$ & $\times$ &$\surd$ \\
        Misclustering rate (for order $K^*$)& - & $\exp(-p^{K/2})$ & $p^{-1}$& $p^{-2}$ &  $\exp(-p^{K/2})$\\
        \hline
    \end{tabular}
    }
    \caption{\small Comparison between previous methods with our method. $^*$We list the result for order-K tensors with $K \geq 3$ and general number of communities $r = \tO(1)$. }\label{tab:comp}
    \vspace{-.2cm}
\end{table*}
\thispagestyle{empty}

{\bf Notation.} We use lower-case letters (e.g., $a,b$) for scalars, lower-case boldface letters (e.g., $\ma,\mtheta$) for vectors, upper-case boldface letters (e.g., $\mX,\mY$) for matrices, and calligraphy letters (e.g., $\tX,\tY$) for tensors of order three or greater. We use $\mone_p$ to denote a vector of length $p$ with all entries to be 1. We use $|\cdot|$ for the cardinality of a set and $\ind\{\cdot\}$ for the indicator function. For an integer $p\in\mathbb{N}_{+}$, we use the shorthand $[p]= \offf{1,2,...,p}$. For a length-$p$ vector $\ma$, we use $a(i)\in\mathbb{R}$ to denote the $i$-th entry of $\ma$, and use $\ma_{I}$ to denote the sub-vector by restricting the indices in the set $I\subset [p]$.  We use  $\onorm{\ma}=\sqrt{\sum_{i}a^2(i)}$ to denote the $\ell_2$-norm, $\onorm{\ma}_1=\sum_i |a_i|$ to denote the $\ell_1$ norm of $\ma$. For two vector $\ma, \mb$ of the same dimension, we denote the angle between $\ma, \mb$ by $\cos \of{\ma, \mb} = \ang{ \ma, \mb} / \onorm{\ma} \onorm{\mb},$ where $\ang{\ma,\mb}$ is the inner product of two vectors and $\cos \of{\ma, \mb} \in [-1,1]$. We make the convention that $\cos \of{\ma, \mb} = \cos \of{\ma^T, \mb^T}$. Let $\tY  \in \bbR^{p_1 \times \cdots \times p_K}$ be an order-$K$ $(p_1,...,p_K)$-dimensional tensor. We use $\tY(i_1,\ldots,i_K)$ to denote the $(i_1,\ldots,i_K)$-th entry of $\tY$. The multilinear multiplication of a tensor $\tS\in \bbR^{r_1\times \cdots \times r_K}$ by matrices $\mM_k \in\mathbb{R}^{p_k\times r_k}$ results in an order-$d$ $(p_1,\ldots,p_K)$-dimensional tensor $\tX$
\[
\tX=\tS \times_1 \mM_1 \times \cdots \times_K \mM_K,
\]
where the entries of $\tX$ are defined by $\tX(i_1,\ldots, i_K)
    =\sum_{(j_1,\ldots,j_K)}\tS(j_1,\ldots,j_K)\mM_1(i_1,j_1)\cdots \mM_K(i_K,j_K).$
% \small
% \begin{align}
%     &\tX(i_1,\ldots, i_K)\\
%     &=\sum_{(j_1,\ldots,j_K)}\tS(j_1,\ldots,j_K)\mM_1(i_1,j_1)\cdots \mM_K(i_K,j_K).
% \end{align} 
% \normalsize
For a matrix $\mY$, we use $\mY_{i:}$ (respectively, $\mY_{:i}$) to denote the $i$-th row (respectively, $i$-th column) of the matrix. Similarly, for an order-3 tensor, we use $\tY_{::i}$ to denote the $i$-th matrix slide of the tensor. We use $\text{Ave}(\cdot)$ to denote the operation of taking averages across elements and $\text{Mat}_k(\cdot)$ to denote the unfolding operation that reshapes the tensor along mode $k$ into a matrix. For a symmetric tensor $\tY\in\mathbb{R}^{p\times \cdots \times p}$, we omit the subscript and use $\text{Mat}(\tY)\in\mathbb{R}^{p\times p^{K-1}}$ to denote the unfolding. For two sequences $\{a_p\}, \{b_p\}$, we denote $a_p\lesssim b_p$ or $a_p=\tO(b_p)$ if $\lim_{p\to\infty}a_p /b_p\leq c$ for some constant $c\geq 0$, $a_p=o(b_p)$ if $\lim_{p\to\infty}a_p/b_p =0$, and $a_p = \Omega(b_p)$ if both $b_p \lesssim a_p$ and $a_p\lesssim b_p$. Throughout the paper, we use the terms ``community'' and ``clusters'' exchangeably.

% {\bf Organization.} The rest of this paper is organized as follows. Section~\ref{sec:model} introduces the degree-corrected tensor block model (dTBM) with three motivating examples and presents the identifiability of dTBM under the angle gap condition. We show the phase transition and the existence of statistical-computational gaps for the higher-order dTBM in Section~\ref{sec:limits}. In Section~\ref{sec:alg}, we provide a polynomial-time two-stage algorithm with misclustering rate guarantees. Numerical studies including the simulation, comparison with other methods, and two real dataset analyses are in Sections~\ref{sec:simulation}-\ref{sec:real}. The main technical ideas we develop for addressing main theorems are provided in Section~\ref{sec:mainproof}. Detailed proofs and extra theoretical results are provided in Appendix.

\section{MODEL FORMULATION}\label{sec:model}

\subsection{dTBM}

Suppose we have an order-$K$ data tensor $\tY \in \bbR^{p \times \cdots \times p}$. For ease of notation, we focus on symmetric tensors in this section; see Supplement~\ref{sec:exten} for the extension to general asymmetric tensors. Assume there exist $r \geq 2$  disjoint communities among the $p$ nodes. We represent the community assignment by a function $z \colon [p]\mapsto[r]$, where $z(i) = a$ for $i$-th node that belongs to the $a$-th community. Then, $z^{-1}(a)=\{i\in[p]\colon z(i)=a\}$ denotes the set of nodes that belong to the $a$-th community, and $|z^{-1}(a)|$ denotes the number of nodes in the $a$-th community. Let $\mtheta=(\theta(1),\ldots,\theta(p))^T$ denote the degree heterogeneity for $p$ nodes. We consider the order-$K$ dTBM~\citep{ke2019community},
\vspace{-0.2cm}
\footnotesize
\begin{equation}\label{eq:model_margin}
    \tY(i_1,\ldots,i_K) = \tS(z(i_1),\ldots,z(i_K)) \prod_{k = 1}^K\theta_{i_k} + \tE(i_1,\ldots,i_K), 
\end{equation}
\normalsize
where $\tS \in \bbR^{r \times \cdots \times r}$ is an order-$K$ tensor collecting the block means among communities, and $\tE\in\mathbb{R}^{p\times \cdots \times p}$ is a noise tensor consisting of independent zero-mean sub-Gaussian entries with variance bounded by $\sigma^2$. The unknown parameters are $z$, $S$, and $\mtheta$. The dTBM can be equivalently written in a compact form of tensor-matrix product:
\begin{equation}\label{eq:model_tensor}
\bbE\tY = \tS \times_1 \mTheta \mM \times_2 \cdots \times_K  \mTheta \mM,
\end{equation}
where $\mTheta = \text{diag}(\theta(1),...,\theta(p)) \in \bbR^{p \times p}$ is a diagonal matrix, $\mM \in \offf{0,1}^{p \times r}$ is the membership matrix associated with community assignment $z$ such that $\mM(i,j)=\ind\{z(i)=j\}$. By definition, each row of $\mM$ has one copy of 1's and 0's elsewhere. Note that the discrete nature of $\mM$ renders our model~\eqref{eq:model_tensor} more challenging than Tucker decomposition. We call a tensor $\tY$ an $r$-block tensor with degree $\mtheta$ if $\tY$ admits~\eqref{eq:model_tensor}. 
%The goal of clustering is to estimate $z$ from a single noisy tensor $\tY$. 
We are particularly interested in high-dimensional regime where $p$ grows whereas $r=\tO(1)$. See Supplement~\ref{sec:examples} for motivated applications of dTBM.

% \begin{rmk}[Comparison with non-degree models]
% Our dTBM uses fewer block parameters than TBM. In particular, every non-degree $r_1$-block tensor can be represented by a \emph{degree-corrected} $r_2$-block tensor with $r_2\leq r_1$. In particular, there exist tensors with $r_1=p$ but $r_2=1$, so the reduction in model complexity can be dramatic from $p$ to 1. This fact highlights the benefits of introducing degree heterogeneity in higher-order clustering tasks.
% \end{rmk}
